---
title: 'Worksheet 1: R Basics'
author: ""
date: ""
---
_This is the first in a series of worksheets for History 8510 at Clemson University. The goal of these worksheets is simple: practice, practice, practice. The worksheet introduces concepts and techniques and includes prompts for you to practice in this interactive document. When you are finished, you should change the author name (above), knit your document to a pdf, and upload it to canvas._

## What is R? 
To start let's define what exactly R is. R is a language and environment for statistical computing and graphics. R provides a variety of statistical and graphical techniques and its very extensible which makes it an ideal language for historians.  

## Foundational Concepts
### Values

There are several kinds of variables in R. Numeric, logical, and strings. 

R takes inputs and returns an output. So for example, if our input is a number the output will be a number. Simply typing a number, below or in the console, will return a number or a **numeric** value. 
```{r}
5
```
Same thing happens if we use a number with a decimal. 
```{r}
5.5483
```
Numbers can be used to do arithmetic. 
```{r}
5 + 5
```

(@) You try, multiple two numbers. 
```{r}
5*55
3*895
```


(@) Can you multiply two numbers and then divide by the result? 
```{r}
275/5
8688/3
35*86/3
```


The next type of value is a **string**. Strings are lines of text. Sometimes these are referred to as character vectors. To create a string, you add text between two quotation marks. For example: 
```{r}
"Go Tigers"
```

(@) Try to create your own string.
```{r}
"Medieval India"
"My name is Megha"
```

You can't add strings using `+` like you can numbers. But there is a function called `paste()` which concatenates character vectors into one. This function is _very_ useful and one you'll use a lot in a variety of circumstances. Here's what that looks like: 

```{r}
paste("Hello", "Clemson Graduate Students")
```

(@) Try it, add two strings together like the above example. 
```{r}
paste("Coding", "is easy")
paste("My", "subject")
```

(@) Can you explain what happened in 2-3 sentences? 
```{r}
We created strings in R using quotation marks and concatenated them using the paste() function. Here, I added two strings together with paste function.
```

The last type are **logical** values like `TRUE` and `FALSE`. (Note that these are all caps.)

```{r}
TRUE
```

```{r}
FALSE
```

These logical values are really useful for testing a statement and comparing values to each other. For example: 
```{r}
3 < 4
5 < 4
6 == 6
```
3 is indeed less than 4 so the return value is `TRUE`. Here are a few more examples:

R also has relational operators and logical operators. Relational operators test how one object relates to another. We've been using some of these above. 

* Equality `==`
* Inequality `!=`
* Less than or Greater than `<` and `>`
* Less than/greater than or equal to `<=` and `>=`

Logical Operators allow you to combine or change the results of comparisons. 
* AND `&`
* OR `|`
* NOT `!`

```{r}
5 == 10
3 < 4
5 != 10
5 != 5
3 == 4 | 3
3 != 4 & 3 != 5
```
(@) Explain, what does the code on each line above do? 
```{r}
5 == 10
3 < 4

5 != 10

5 != 5

3 == 4 | 3

3 != 4 & 3 != 5

these lines of code are performing logical comparisons in R, demonstrating how equality and inequality operators work and how logical OR is interpreted when used with numeric values.
```

(@) Create your own comparison. 
```{r}
2==2

3 != 3

3 * 2 == 6

4+9 != 13



---

### Variables

Values are great but they are made so much more powerful when combined with **Variables** which are a crucial building block of any programming language. Simply put, a variable stores a value. For example if I want x to equal 5 I can do that like this: 
```{r}
x <- 5
```
`<-` is known as an assignment operator. Technically, you could use `=:` here too but it is considered bad practice and can cause complicated issues when you write more advanced code. So its important to stick with `<-` whenever you are coding in R. 

I can also add a string or character vector to a variable. 
```{r}
x <- "Go Tigers!"
```

Variable names can be almost anything. 
```{r}
MyFavoriteNumber <- 25
```
Variable or object names must start with a letter, and can only contain letters, numbers, `_`, and `.`. You want your object names to be descriptive, so you’ll need a convention for multiple words. People do it different ways. I tend to use periods but there are several options: 

```
i_use_snake_case 
otherPeopleUseCamelCase 
some.people.use.periods 
And_aFew.People_RENOUNCEconvention
```

Whatever you prefer, be consistent and your future self will thank you when your code gets more complex. 

(@) You try, create a variable and assign it a number: 

```{r}
 my_number <- 10
```

(@) Can you assign a string to a variable?

```{r}
> my_string <- "Hello, World!"
```

Once we've assigned a variable we can use that variable to run calculations just like we did with raw numbers.
```{r}
x <- 25
x * 5
```

R can also handle more complex equations. 
```{r}
(x + x)/10
(x + x * x) - 100
```

And we could store the output of a calculation in a new variable:
```{r}
My.Calculation <- (x + x)*10
```

(@) You try. Assign a number to x and a number to y. Add those two numbers together.
```{r}
(x + y)
[1] 11
```

(@) Can you take `x` and `y` and multiply the result by 5? 
```{r}
> (x + y)*5
[1] 55
```

(@) Try creating two variables with names other than `x ` and `y`. Descriptive names tend to be more useful. Can you multiply the contents of your variables? 
```{r}
> first_number * second_number
[1] 50
```

(@) Try creating two variables that store strings. Can you concatenate those two variables together? 
```{r}
> first_word <- "Digital"
> second_word <- "History"
> paste(first_word, second_word, sep="-")
[1] "Digital-History"
```

### Vectors

If we have a lot of code and rely on just variables, we're going to have a lot of variables. That's where **vectors** come into play. Vectors allow you to store multiple values. All variables in R are actually already vectors. That's why when R prints an output, there is a `[1]` before it. That means there is one item in that vector. 

```{r}
myvalue <- "George Washington"
myvalue
```
In this instance "George Washington" is the only item in the variable myvalue. But we could add more. To do that we use the `c()` function which combines values into a vector. 
```{r}
myvalue <- c("George Washington", "Franklin Roosevelt", "John Adams")
myvalue
```
You'll notice that the output still only shows `[1]` but that doesn't mean there is only one item in the list. It simply means George Washington is the first. If we use `length()` we can determine the number of items in this vector list. 
```{r}
length(myvalue)
```
We could get the value of the 2nd or 3rd item in that list like this: 
```{r}
myvalue[2]
```
We could also create a vector of numbers:
```{r}
my.numbers <- c(2, 4, 6, 8)
```

And we could then do calculations based on these values: 
```{r}
my.numbers * 2
```
(@) Explain in a few sentences, what happened in the code above?

>
```{r}
my.numbers <- c(2, 4, 6, 8)
my.numbers * 2
[1]  4  8 12 16
the operation would multiply each element of the vector i.e, 2, 4, 6, 8 by 2, resulting in a new vector with doubled values: 4, 8, 12, and 16.
```

Lets try something slightly different. 
```{r}
my.numbers[3] * 2
```
(@) Explain in a few sentences, what happened in the code above? 

>
```{r}
my.numbers[3] * 2
[1] 12
The [3] indicates that we are interested in the third element of this vector.
The third element) is then multiplied by 2.
The result is 12.
```


(@) You try, create a list of five items and store it in a descriptive variable. 
```{r}
MagicalItemsList <- list("Crystal Ball", "Wizard's Wand", "Enchanted Cloak", "Potion of Wisdom", "Book of Spells")
```


### Built In Functions

R also has **built in functions**. We've already used a couple of these: `paste()` and `c()`. But there are others, like `sqrt()` which does what you think it does, finds the square root of a number. 
```{r}
sqrt(1000)
```

Many functions have options that can be added to them. For example, the `round()` function allows you to include an option specifying how many digits to round to. 

You can run it without that option and it'll use the default: 
```{r}
round(15.492827349)
```

Or we can tell it to round it to 2 decimal places.
```{r}
round(15.492827349, digits = 2)
```

How would you know what options are available for each function in R? Every function and package in R comes with **documentation** or a **manual** that is built into R studio and can be pulled by by typing a question mark in front of the function in your console. These packages will commonly give you examples of how to use the function and syntax for doing so. They are incredibly useful. 

We can ask R to pull up the documentation like this:
```{r}
?round()
```

(@) Now you try, find the documentation for the function  `signif()`.
```{r}
signif(123.456789, digits = 3)
[1] 123
```

In real life, you typically you wouldn't want to store this code in your script file. You probably don't need to pull up the documentation for the function every time you run that piece of code. But for the purposes of this worksheet we're adding it to our `.Rmd` document. 

(@) Use the console to find the documentation for `floor()`? Try it and then tell me, what does that function do? 

>
```{r}
The floor() function in R is used to round down numerical values to the nearest integer. For example, I used the the following to help me understand -
floor(3.8)
floor(0.0001)
```

### Data Frames & Packages

R is the language of choice for most data scientists and that is because of its powerful suite of data analysis tools. Some are built into R, like `floor()` which we looked at above. But others come from packages that you have to install. Most programming languages have some sort of package system although every language calls it a slightly different thing. Ruby has gems, python has eggs, and php has libraries. In R these are called packages or libraries and they are hosted by the Comprehensive R Archive Network or more commonly, CRAN. CRAN is a network of ftp and web servers around the world that store identical, up-to-date, versions of code and documentation for R. You didn't know it but you already used CRAN when you looked up the documentation for `floor()` above. If you go to the CRAN webpage and look at the [list of available R packages](https://cran.r-project.org/web/packages/available_packages_by_name.html) you'll see just how many there are!

So there are many packages but there are also some that are indispensable. You'll use some packages over and over for this class. We'll get into some of those in the next two worksheets but for now lets look at one basic package for data. 

Lets start with the `tibble()` package which allows you to create and work with **data frames**. What is a data frame? Think of it as a spreadsheet. Each column can contain data - including numbers, strings, and logical values. 

Lets load the `tibble()` package. If you have this package installed this line of code will work. If not, you'll get an error that says something like: `## Error in library("tibble"): there is no package called 'tibble'`. If that's the case (it probably is), then no worries - we can install it. To install a package run `install.packages("tibble")` in your console and R will download and install the package. (Remember this isn't code that we want or need to store in our document, so run it in your console not in the `.Rmd` document.)
```{r}
library(tibble)
```

The tibble package will allow you to create dataframes. Lets use some built in data in R to demonstrate this. This dataset contains the measurements in centimeters of sepal and petal width and length. Lets start by figuring out what kind of data iris is: 
```{r}
typeof(iris)
```

We can see that iris is a list. But we want to use the tibble package to turn it into a dataframe. Thats very easy, we can simply run: 
```{r}
as_tibble(iris)
```

That is useful. But we can also create our own data frame from vectors. That means we can create the values in each field. For example: 

```{r}
tibble(
x = 1:5, 
y = 1, 
z = x * 2 + y
)
```
What is going on here? Each line within this function (x,y,z) creates a new column and determines the values in each row. So the first line creates five rows with values sequentially from 1 to 5. The second column contains only the value 1 in all five rows. 

(@) What is going on in row z? Can you explain the logic here?

>
```{r}
tibble(
+     x = 1:5, 
+     y = 1, 
+     z = x * 2 + y
+ )
# A tibble: 5 × 3
      x     y     z
  <int> <dbl> <dbl>
1     1     1     3
2     2     1     5
3     3     1     7
4     4     1     9
5     5     1    11
Since x ranges from 1 to 5 and y is always 1, the calculation for each row in z will be as follows:
For the first row: z = 1 * 2 + 1 = 3
For the second row: z = 2 * 2 + 1 = 5
For the third row: z = 3 * 2 + 1 = 7
For the fourth row: z = 4 * 2 + 1 = 9
For the fifth row: z = 5 * 2 + 1 = 11

```


## Loading Dataframes from Packages

Lets also install the data package that I've created for our class. This package contains a variety of historical datasets that you can use to complete your assignments this semester. However, this package is not on CRAN. That's okay. We can still install it from GitHub. To do that we'll need to use the `devtools` package. We'll install devtools and then use it to install our class's package which is hosted on GitHub. 

First, install the `devtools` package.

Our class's package is called `DigitalMethodsData` and its hosted on [GitHub](https://github.com/regan008/DigitalMethodsData). Use the directions on the repository page and what you learned about packages above to install this package. 

After installing the package, we should load it: 
```{r}
library(DigitalMethodsData)
```

What kind of datasets are available in this package? We can use the help documentation to find out. 

Run `help(package="DigitalMethodsData")` in your console to pull up a list of datasets included in this package. 

For demonstration purposes we'll use the `gayguides` data here. Pull up the help documentation for this dataset. What does the documentation tell you abou the scope of this dataset? 

>

To use a dataset included in this package we first need to load it. We can do so like this:

```{r}
data(gayguides)
```

Notice that you now have a loaded dataset in your environment pane. It shows us that there are 60,698 observations (rows) of data and 14 variables (columns).

Let's now look at our data. You can use the `head()` function to return the first part of an object or dataset.
```{r}
head(gayguides)
```
That gives us the first 6 rows of data. Its really useful if you just want to peak into the dataset but don't want to print out all 60k+ rows.

(@) The default is to print six rows of data. Can you modify the above code to print out the first 10 rows? 
```{r}
> help(package="DigitalMethodsData")
> head(gayguides, 10)
X   ID        title                        description streetaddress           type
1   1 3213       'B.A.'                    (woods & ponds)               Cruising Areas
2   2 2265 'B.A.' Beach                           2 mi. E.        Rte. 2 Cruising Areas
3   3 3269 'B.A.' Beach                 nr. Salt Air Beach               Cruising Areas
4   4 3388 'B.A.' Beach      nr. Evergreen Floating Bridge               Cruising Areas
5   5 3508 'B.A.' Beach               (2 mi. E. on Rte. 2)               Cruising Areas
6   6 5116 'B.A.' Beach      nr. Evergreen Floating Bridge               Cruising Areas
7   7  780 'B.A.' Beach on Russian River, at Wohler Bridge               Cruising Areas
8   8 3810 'B.A.' Beach                           2 mi. E.        Rte. 2 Cruising Areas
9   9 4979 'B.A.' Beach                 nr. Salt Air Beach               Cruising Areas
10 10 5191 'B.A.' Beach      nr. Evergreen Floating Bridge               Cruising Areas
amenityfeatures           city state Year notes      lat        lon
1      Cruisy Area    Lake Placid    NY 1982       44.27949  -73.97987
2      Cruisy Area           Troy    NY 1981       42.72841  -73.69178
3      Cruisy Area Salt Lake City    UT 1981       40.74781 -112.18727
4      Cruisy Area        Seattle    WA 1981       47.60621 -122.33207
5      Cruisy Area           Troy    NY 1982       42.72841  -73.69178
6      Cruisy Area        Seattle    WA 1983       47.60621 -122.33207
7      Cruisy Area  Russian River    CA 1983       38.91163 -123.01105
8      Cruisy Area           Troy    NY 1983       42.72841  -73.69178
9      Cruisy Area Salt Lake City    UT 1983       40.74781 -112.18727
10     Cruisy Area        Seattle    WA 1984       47.60621 -122.33207
status
1  Location could not be verified. General city or location coordinates used.
2  Location could not be verified. General city or location coordinates used.
3  Location could not be verified. General city or location coordinates used.
4  Location could not be verified. General city or location coordinates used.
5  Location could not be verified. General city or location coordinates used.
6  Location could not be verified. General city or location coordinates used.
7  Location could not be verified. General city or location coordinates used.
8  Location could not be verified. General city or location coordinates used.
9  Location could not be verified. General city or location coordinates used.
10 Location could not be verified. General city or location coordinates used.

```

(@) Can you find the last 6 rows of data? 
```{r}
tail(gayguides, 6)
X    ID           title               description streetaddress       type
60693 60693 PR351       Boccachio            (Natives only)               Bars/Clubs
60694 60694 PR366       Boccachio            (Natives only)               Bars/Clubs
60695 60695 PR377       Boccachio            (Natives only)               Bars/Clubs
60696 60696 GU002   Mi Elena Club (Inquire locally) (Agana)          <NA> Bars/Clubs
60697 60697 GU015 Star Cafe & Bar Downtown, opp. Agana Loop          <NA> Bars/Clubs
60698 60698 GU019 Star Cafe & Bar Downtown, opp. Agana Loop          <NA> Bars/Clubs
amenityfeatures
60693                                                                                     (G),(*)
60694                                                                                         (G)
60695                                                                                         (L)
60696                                                         (M) - Mixed Crowds - Some Straights
60697 (M) - Mixed Crowds - Some Straights,(YC) - Young and/or collegiate types,(*) - Very popular
60698 (M) - Mixed Crowds - Some Straights,(YC) - Young and/or collegiate types,(*) - Very popular
city state Year notes      lat       lon
60693 Hato Ray    PR 1977       18.42258 -66.05096
60694 Hato Ray    PR 1979       18.42258 -66.05096
60695 Hato Ray    PR 1980       18.42258 -66.05096
60696    Agana    GU 1980  <NA> 13.47628 144.75022
60697    Agana    GU 1975  <NA> 13.47628 144.75022
60698    Agana    GU 1974  <NA> 13.47628 144.75022
status
60693 Location could not be verified. General city or location coordinates used.
60694 Location could not be verified. General city or location coordinates used.
60695 Location could not be verified. General city or location coordinates used.
60696 Location could not be verified. General city or location coordinates used.
60697 Location could not be verified. General city or location coordinates used.
60698 Location could not be verified. General city or location coordinates used.
```

(@) Reflect on the previous two prompts. How did you figure these out? 

>
```{r}
The key to figuring these out lies in understanding what the head() and tail() functions do in R. head() is used to display the first few rows of a dataset, and tail() for the last few rows.

```

The `str()` function is another very useful function for understanding a dataset. It compactly displays the internal structure of an R object. 
```{r}
str(gayguides)
```
There is a ton of useful info here. First, we see that this is a data.frame and that it has 60698 obs. of  14 variables. Second, we see the names of all the variables (columns) in the dataset. Lastly, it shows us what type of data is contained in each variable. For example, the variable `state` is a chr or character vector while `Year` (note the capitalization) is numeric. 

The `$` operator allows us to get a segment of the data. 
```{r}
head(gayguides$title)
```
While that's useful, this particular dataset happens to be sorted by title. So the first rows are all labeled some version of B.A. beach. That's not very useful. What if we want to see the 500th item in this list? Well, we can pull that up like this: 
```{r}
gayguides$title[500]
```

(@) Break this down. Whats going on here? What do each of the elements of this code mean? (gayguides, $, title, and [500]) 

>
```{r}
gayguides$title[500] extracts the 500th title from the title column in the gayguides data frame. 
```

(@) The Year variable contains the year (numeric) of each entry. Can you find the earliest year in this dataset? (Using code, don't cheat and use the documentation!)
```{r}
min(gayguides$Year, na.rm = TRUE)
[1] 1965
```


(@) How about the latest year? 
```{r}
max(gayguides$Year, na.rm = TRUE)
[1] 1985
```


(@) Add another dataset from `DigitalMethodsData`. How many observations are included in this dataset? Run through some of the examples from above to learn about the dataset.
```{r}
help(package="DigitalMethodsData")
> data("almshouse_admissions")
> head(almshouse_admissions, 10)
Last.Name First.Name Year.Admitted Month.Admitted Gender Descriptions.by.Clerk
1      ABBOT     ROSINA          1803              7      1                    10
2       ABEL       JANE          1801              1      1                    50
3       ABEL       MARY          1801              1      1                     6
4   ABERVINE       JOHN          1800             12      0                     3
5   ACHILLER      KITTY          1802              1      1                     6
6       ADAM     HANNAH          1801              8      1                     0
7      ADAMS        ANN          1803              3      1                     5
8      ADAMS   ANN MARY          1803              3      1                     6
9      ADAMS    FRANCIS          1803             11      0                     6
10     ADAMS      JACOB          1800             11      0                     2
Reason.Discharged YearDischarged MonthDischarged
1                  2           1804               1
2                  4           1801               8
3                  3           1801               6
4                  1           1801               2
5                  3           1802               1
6                  1           1802               5
7                  1           1803               5
8                  1           1803               5
9                  3           1803              12
10                 1           1801               4
> tail(almshouse_admissions, 6)
Last.Name First.Name Year.Admitted Month.Admitted Gender Descriptions.by.Clerk
1395       NEGRO       ADAM          1802              5      0                     1
1396       NEGRO       AMBO          1802              5      1                     1
1397   black boy      BANAM          1803              1      0                    52
1398   black boy      BANAM          1803              1      0                     1
1399 black child    CHARLES          1801              3      0                    52
1400     mullato       CATO          1800              7      0                    19
Reason.Discharged YearDischarged MonthDischarged
1395                 2           1802               7
1396                 2           1802               5
1397                 3           1803               1
1398                 4           1803               1
1399                 0              1               0
1400                 1           1804               1
```

(@) What did you learn about the dataset? What can you tell me about it? 
```{r}
After running these commands, I learnt about dataset, such as the number of observations, the types of variables it contains, and a glimpse of the data. This information helped me understand the nature of the dataset, what kind of information it contains, and potentially how it can be used for research. I also learnt about so many features of R.
```

CONGRATS! You've completed your first worksheet for Digital Methods II. That wasn't so bad, right? 

















